## Reproducible Research first assignment

# Preparation

For the data analysis the package 'reshape2' is used. The following code checkes whether it is installed, installs it if not and loads the package

```{r cache=TRUE}
if (library("reshape2", logical.return = TRUE) == FALSE) {
    install.packages("reshape2")
}
library("reshape2")
```

# 1. Read in the data

The zip file with the dataset is extracted into the home directory and then read in

```{r cache=TRUE}
unzip("RepData_PeerAssessment1/activity.zip")
activity <- read.csv("activity.csv", header = TRUE, nrows = 17570, colClasses = c("numeric",
    "Date", "numeric"))
```

# 2. What is the mean total number of steps taken each day?

Reshape the dataset to get the total numbers of steps per day


```{r cache=TRUE, fig.width=11, fig.height=6}
myMelt <- melt(activity, id = "date")
myTotal <- dcast(myMelt, date ~ variable, sum)
Create the histeogram of total numbers of steps taken each day

hist(myTotal$steps, main = "Total number of steps")
plot of chunk unnamed-chunk-4
```

Calculate the the mean and median of total numbers of steps taken each day, NAs need to be removed as otherwise the mean and median are NA

```{r cache=TRUE}
mean(myTotal$steps, na.rm = TRUE)
## [1] 10766
median(myTotal$steps, na.rm = TRUE)
## [1] 10765
```

# 3. What is the average daily activity pattern?

Reshape the data to get the mean for each interval

```{r cache=TRUE, fig.width=11, fig.height=6}
myMelt <- melt(activity, id = "interval", na.rm = TRUE)
myInterval <- dcast(myMelt, interval ~ variable, mean)
Create a time series plot of the means for each interval

plot(myInterval$interval, myInterval$steps, type = "l")
plot of chunk unnamed-chunk-7
```

Calculate the interval with the maximum average of steps

```{r cache=TRUE}
myInterval[myInterval$steps == max(myInterval$steps), 1]
## [1] 835
```

# 4. Imputing missing values

Computing the number of missing cases

```{r cache=TRUE, fig.width=11, fig.height=6}
nrow(activity[!complete.cases(activity), ])
## [1] 2304
Creating a new dataset in which the missing values for steps are replaced with the mean of the respective interval

x <- rep(myInterval$steps, 61)
activityNew <- activity
activityNew$steps[is.na(activityNew$steps)] <- x[is.na(activityNew$steps)]
Reshape the new dataset to get the total numbers of steps per day

myMeltNew <- melt(activityNew, id = "date")
myTotalNew <- dcast(myMeltNew, date ~ variable, sum)
Creating the histeogram for the new dataset

hist(myTotalNew$steps, main = "Total number of steps")
plot of chunk unnamed-chunk-12
```

Calculating mean and median for the new dataset

```{r cache=TRUE}
mean(myTotalNew$steps)
## [1] 10766
median(myTotalNew$steps)
## [1] 10766
```

As one can see the mean and median above differ slightly from the mean ( 1.0766 × 104 ) and median ( 1.0765 × 104 ) of the dataset without replacing the NAs When replacing the NAs with the average of the respective interval the median and mean have the same value, indicating a more even distribution.

# 5. Are there differences in activity patterns between weekdays and weekends?

Creating an new factor variable called 'Days' with the levels weekday and weekend

```{r cache=TRUE, fig.width=11, fig.height=6}
x <- as.factor(weekdays(activityNew$date))
levels(x) <- list(Weekday = "Friday", Weekday = "Monday", Weekend = "Saturday",
    Weekend = "Sunday", Weekday = "Thursday", Weekday = "Tuesday", Weekday = "Wednesday")
activityNew$Days <- x
Preparing the dataset to create a time series plot of the means for each interval for both weekend and weekday by subsetting and reshaping

myDaySub <- activityNew[activityNew$Days == "Weekday", c("interval", "steps")]
myEndSub <- activityNew[activityNew$Days == "Weekend", c("interval", "steps")]

myMeltDay <- melt(myDaySub, id = "interval", na.rm = TRUE)
myIntervalDay <- dcast(myMeltDay, interval ~ variable, mean)

myMeltEnd <- melt(myEndSub, id = "interval", na.rm = TRUE)
myIntervalEnd <- dcast(myMeltEnd, interval ~ variable, mean)
Creating a time series plot of the means for each interval for both weekend and weekday

par(mfrow = c(2, 1), mar = c(4, 4, 1, 1))
plot(myIntervalDay$interval, myIntervalDay$steps, type = "l", col = "blue",
    main = "Weekday", ylab = "Number of Steps", xlab = "", xaxt = "n", ylim = c(0,
        250))
plot(myIntervalEnd$interval, myIntervalEnd$steps, type = "l", col = "blue",
    main = "Weekend", ylab = "Number of Steps", xlab = "interval", ylim = c(0,
        250))
plot of chunk unnamed-chunk-16
```

As can be seen the average steps taken on the weekend are much more distributed ver the whole day, while during the weekdays the average spikes at interval 835